Mixture Models for Single-Cell Assays with Application to Vaccine Studies

### 2013-05-23 MIMOSA.tex
More revisions to text

A motivating example from vaccine research is the flow cytometric intracellular cytokine staining (ICS) assay, which is used to identify and quantify subjects' immune responses to a vaccine. Upon vaccination, antigen in the vaccine is taken up and presented to CD4 or CD8 T-cells via antigen presenting cells.
 While not all T-cells can recognize all antigens, those that recognize antigens in the vaccine become \emph{activated} and produce a variety of cytokines, further promoting the immune response.
 After activation, this antigen-specific subpopulation proliferates and can persist in the immune system for some time providing \emph{memory} that can more rapidly recognize the same antigen again in the future~\citep{McKinstry:2010ei}. The antigen--specific T-cell subpopulations (i.e. the subset that can respond to one specific antigen) constitute a very small fraction of the total number of CD4 and CD8 T-cells. The ICS assay measures the number of antigen-specific T-cells in PBMC or whole blood by measuring cytokine production in response to activation following stimulation by an antigen that closely matches what was present in the original vaccine.
Individual cells are labelled using fluorescently conjugated antibodies against cell--surface biomarkers (CD3, CD4, and CD8), used to subset T-cells, and functional biomarkers (cytokines) used to define antigen specific T-cells~\citep{Horton:2007tsa,DeRosa:2004wp,Betts:2006dw}.
A sufficiently large number of cells must be collected (on the order of 50,000 to 100,000 T--cells) to ensure that the rare cell populations can be detected.
Subsequently, each individual cell is classified as either positive or negative for each maker based on predetermined thresholds, then the number of cells matching each subpopulation phenotype is counted.

These counts are compared between antigen stimulated and unstimulated samples from a subject to identify significant differences. Subjects who generate a response after stimulation are called \textit{responders}, whereas subjects that do not show any differences are called \textit{non-responders}. The comparisons between stimulated and unstimulated samples are typically done within a time point (\textit{i.e.,} within pre--vaccination and within post--vaccination time points), rather than directly between pre--and post--vaccine time points to avoid confounding vaccine effect with date or time of blood draw; additionally not all vaccine trials collect baseline (pre--vaccination) samples.
In many immunological studies, the size of the functionally distinct subpopulations (\textit{i.e.}, the number of positive cells) is very low (on the order of 10s of cells, with effect sizes on the order of $10^{-4}$, relative to the total number of cells), and real biological differences might be difficult to detect.

@@

The framework developed in this paper, named MIMOSA (Mixture Models for Single Cell Assays), addresses these issues explicitly. In our model, cell counts are modelled by a binomial (or multinomial in the multivariate case) distribution and information is shared across subjects by means of a prior distribution placed on the unknown proportion(s) of the binomial (or multinomial) likelihood.
In order to discriminate between responders and non-responders, the prior is written as a mixture of two beta (or Dirichlet in the multivariate case) distributions where the hyper-parameters for each mixture component are shared across subjects.
This sharing of information helps regularize proportion estimates when the cell counts are small, which is typical with single-cell assays, and increases sensitivity and specificity when detecting responders.
Because our framework is multivariate in nature, multiple cell subsets can be modelled simultaneously, which could help detect small biological changes that are spread out across multiple cell subsets~\citep{Nason:2006dx}. Our paper is organized as follows; Section \ref{s:data} introduces the data and notations used in the paper. In Section \ref{s:DEone}, we present our model for testing differential biomarker expression in the univariate case. Section \ref{s:results} compares our approach to alternative methods and tests the robustness of our model. In Section \ref{s:demarkercombos} we present a multivariate extension of our model that can be used to test multivariate biomarker differential expression and present some results using a single-cell gene expression data. Finally, in Section \ref{s:discussion} we discuss our findings and future work.

@@

\textit{Flow cytometry}:
The primary dataset used here is an ICS data set generated as part of a  trial testing the GeoVax DNA and MVA (Modified Vaccinia Ankara) HIV vaccine in a prime-boost regimen (prime at zero and two months, boost at four and six months)~\citep{Goepfert:2011ci}.
The goal of this data set was to assess the immune response to the vaccine across multiple antigen stimulations, time points, cytokines and T-cell subsets.
Here, we analyze a subset of the data consisting of 98 subjects from the vaccine group at two time points: day 0 and day 182.
Three cytokines (IFN-$\gamma$, TNF$\alpha$ and IL2) were measured at the single-cell level for each subject and time point, with and without stimulations with an antigen (here we focus on HIV Envelope peptide pool) matching part of the vaccine. For ease of presentation we restricted ourselves to the CD4+ T-cell subsets. Samples on day 0 were taken just before vaccination and no response is expected there. The corresponding samples can be used as negative controls. Conversely, day 182 (26 weeks) should be close to the immunogenicity peak, and many subjects are expected to respond, for some cytokines at least. In this data set a median of 51K and 58K T--cells were collected for the stimulated and unstimulated samples, respectively (IQR $\approx$ 37K and 43K, respectively) were collected for the stimulated and unstimulated samples. The effect sizes (differences in the proportions of stimulated and unstimulated cells) are on the order of $10^{-4}$, with the number of positive cells typically ranging from 1--70, with a few samples exhibiting large numbers of positive events.

>>>>

A motivating example from vaccine research is the flow cytometric intracellular cytokine staining (ICS) assay, which is used to identify and quantify subjects' immune responses to a vaccine. Upon vaccination, antigen in the vaccine is taken up and presented to CD4 or CD8 T-cells via antigen presenting cells.
 While not all T-cells can recognize all antigens, those that recognize antigens in the vaccine become \emph{activated} and produce a variety of cytokines, further promoting the immune response.
 After activation, this antigen-specific subpopulation proliferates and can persist in the immune system for some time providing \emph{memory} that can more rapidly recognize the same antigen again in the future~\citep{McKinstry:2010ei}. The antigen--specific T-cell subpopulations (i.e. the subset that can respond to one specific antigen) constitute a very small fraction of the total number of CD4 and CD8 T-cells (on the order of 0.1\% for HIV vaccine studies like the one shown here). The ICS assay measures the number of antigen-specific T-cells in PBMC or whole blood by measuring cytokine production in response to activation following stimulation by an antigen that closely matches what was present in the original vaccine.
Individual cells are labelled using fluorescently conjugated antibodies against cell--surface biomarkers (CD3, CD4, and CD8), used to subset T--cells, and functional biomarkers (cytokines) used to define antigen specific T--cells~\citep{Horton:2007tsa,DeRosa:2004wp,Betts:2006dw}.
A sufficiently large number of cells must be collected (on the order of 50,000 to 100,000 T--cells) to ensure that the rare cell populations can be detected.
Subsequently, each individual cell is classified as either positive or negative for each maker based on predetermined thresholds, then the number of cells matching each subpopulation phenotype is counted.

These counts are compared between antigen stimulated and unstimulated samples from a subject to identify significant differences. Subjects who generate a response after stimulation are called \textit{responders}, whereas subjects that do not show any differences are called \textit{non-responders}. Vaccine--specific response is determined when a subject has no response at a pre--vaccine time point and a response at a post--vaccine time point, and furthermore, when placebo recipients exhibit no response at either time point. The comparisons between stimulated and unstimulated samples are typically done within a time point rather than directly between pre--and post--vaccine time points to avoid confounding vaccine effect with date or time of blood draw, and additionally because baseline (pre--vaccination) samples are not always available.
In many immunological studies, the size of the functionally distinct subpopulations (\textit{i.e.}, the number of positive cells) is very low (on the order of 10s of cells, with effect sizes on the order of $10^{-4}$, relative to the total number of cells), and real biological differences might be difficult to detect.

@@

The framework developed in this paper, named MIMOSA (Mixture Models for Single Cell Assays), addresses these issues explicitly. In our model, cell counts are modeled by a binomial (or multinomial in the multivariate case) distribution and information is shared across subjects by means of a prior distribution placed on the unknown proportion(s) of the binomial (or multinomial) likelihood.
In order to discriminate between responders and non-responders, the prior is written as a mixture of two beta (or Dirichlet) distributions where the hyper-parameters for each mixture component are shared across subjects.
This sharing of information helps regularize proportion estimates when the cell counts are small, which is typical with the single-cell assays considered here, and increases sensitivity and specificity when detecting responders.
Because our framework is multivariate in nature, multiple cell subsets can be modeled simultaneously, which could help detect small biological changes that are spread out across multiple cell subsets~\citep{Nason:2006dx}. Our paper is organized as follows; Section \ref{s:data} introduces the data and notations used in the paper. In Section \ref{s:DEone}, we present our model for testing differential biomarker expression in the univariate case. Section \ref{s:results} compares our approach to alternative methods and tests the robustness of our model. In Section \ref{s:demarkercombos} we present a multivariate extension of our model that can be used to test multivariate biomarker differential expression and present some results using a single-cell gene expression data. Finally, in Section \ref{s:discussion} we discuss our findings and future work.

@@

\textit{Flow cytometry}:
The primary dataset used here is an ICS data set generated as part of a  trial testing the GeoVax DNA and MVA (Modified Vaccinia Ankara) HIV vaccine in a prime-boost regimen (prime at zero and two months, boost at four and six months)~\citep{Goepfert:2011ci}.
The goal of this data set was to assess the immune response to the vaccine across multiple antigen stimulations, time points, cytokines and T-cell subsets.
Here, we analyze a subset of the data consisting of 98 subjects from the vaccine group at two time points: day 0 and day 182.
Three cytokines (IFN-$\gamma$, TNF$\alpha$ and IL2) were measured at the single-cell level for each subject and time point, with and without stimulations with an antigen (here we focus on HIV Envelope peptide pool) matching part of the vaccine. For ease of presentation we restricted ourselves to the CD4+ T-cell subsets. Samples on day 0 were taken just before vaccination and no response is expected there. The corresponding samples can be used as negative controls. Conversely, day 182 (26 weeks) should be close to the immunogenicity peak, and many subjects are expected to respond, for some cytokines at least. In this data set a median of 51K and 58K T--cells were collected for the stimulated and unstimulated samples, respectively (IQR $\approx$ 37K and 43K, respectively). The effect sizes (differences in the proportions of stimulated and unstimulated cells) are on the order of $10^{-4}$, with the number of positive cells typically ranging from 1--70, with a few samples exhibiting large numbers of positive events.
